---
title: "Find Highly Variable Genes"
output:
  html_document:
    df_print: paged
    toc: yes
    number_sections: true
    toc_float: true
    self_contained: true
    mathjax: default
    code_download: true
---







# Load data
On the CITEseq data, we play around with HVGs (highly variable genes), i.e.
superpoissonian genes.

```{r loadCITE}
file <- "~/sds/sd17l002/p/scRNAseq_datasets/CITEseq_NatMethods_2017/data/GSE100866_CBMC_8K_13AB_10X-RNA_umi.csv.gz"
rawC <- as.matrix(read.csv(gzfile(file), row.names = 1))

is_mouse <- colSums(rawC[grepl("MOUSE", rownames(rawC)),]) / colSums(rawC) > .1
rawC <- rawC[grepl("HUMAN", rownames(rawC)), ! is_mouse]

```



# ColSums, Gene Means and Poisson raw counts

```{r}
cs <- colSums(rawC)

# estimate each gene's true mean by normalizing with library size first:
normC <- t( t(rawC) / cs)
allM <- apply(normC, 1, mean)
allV <- apply(normC, 1, var)


# simulate poisson raw counts for our data:
pCounts <- t(sapply(allM, function(mu) rpois(ncol(rawC), lambda = mu * cs)))
rownames(pCounts) <- paste0("poisson_", 1:nrow(pCounts))
colnames(pCounts) <- paste0("Cell_", 1:ncol(pCounts))
```



# [code chunks]

```{r}
normP <- t( t(poiss) / colSums(rawC))
allMP <- apply(normP, 1, mean)
allVP <- apply(normP, 1, var)
plot(allM, allV/allM, log = "xy", pch = 20, cex=.1, col = "#00000040")
points(allMP, allVP / allMP, col = "red", pch = 20, cex=.1)
```

# Local quantile fit

To select HVGs, we want to know where the Poisson-regime ends. For this, we'll
fit lines through the mean/median and the 90%-quantile using local quantile regression:
```{r}
# Simon wants to filter out low-abundance genes. A very conservative cutoff is
# the mean resulting from 1 small cell (nUMI = 500) having an UMI of 1, the rest 0.
threshold <- mean(c(1/500,  rep(0, ncol(rawC)-1)))

# ease of typing (Poisson stats):
is_na <-  is.na(allVP/allMP) | is.na(allMP)
 mu <- allMP[!is_na & allMP > threshold]
vmr <- allVP[!is_na & allMP > threshold] / allMP[!is_na & allMP > threshold]

x <- log(mu)
y <- log(vmr)


 # going over all x takes too long:
xrange <- log(c(max(1e-9, min(allM), min(allMP)), max(max(allM), max(allMP))))
bin_n <- 75
bin_x <- seq(xrange[1], xrange[2], length.out = bin_n)


library(quantreg)
mm <- cbind( 1, x, x^2 )
yfit <-
 sapply( bin_x, function(xp) {
   fit <- rq.wfit( mm, y, .9, dnorm( x, mean=xp, sd=2 ) )
   fit_median <- rq.wfit( mm, y, .5, dnorm( x, mean=xp, sd=2 ) )
   c(   q90 = fit$coefficients %*% c( 1, xp, xp^2 ),
     median = fit_median$coefficients %*% c( 1, xp, xp^2 ))
   })


plot(log(allM), log(allV/allM), pch=20, cex=.1)
points(x, y, pch=20, cex=.1, col ="red")
lines( bin_x, yfit[1,], lwd = 2, col="orange" )
lines( bin_x, yfit[2,], lwd = 2, col="magenta" )
lines( bin_x, yfit["median", ] + 10*(yfit["q90", ] - yfit["median",]), lwd = 2, lty = "dashed", col="green" )

```
The green line is at the moment a bit arbitrarily chosen as the median plus
10 times the difference from median to 90 quantile. We'll decide later what works
best.










# End of Script

```{r}
devtools::session_info()
```


